---
title: "Plotting LapTime for Formula 1 2022 World Competition"
author: "Luan Vieira"
output:  
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data Prep

Loading libraries

```{r,message=FALSE}
library(ggplot2)
library(gridExtra)
library(ggpmisc)
library(readxl)
library(dplyr)
```

Loading data

```{r,message=FALSE}
df = read_excel("./files/Races_cleaned.xlsx")
df = df %>% rename('Local' = "PlaceBrazilianPortuguese")
```

Formating millisecond to friendlier format.

```{r}
timeHMS_formatter <- function(x) {
  x[is.na(x)] <- 0
#  h <- trunc(x / 24 / 60 / 1000)
  m <- trunc(x / 60 / 1000)
  s <- (x / 1000) %% 60
  s <- round(s, 2)
  lab <- sprintf('%02dmin:%02ds', m, s) # Format the strings as HH:MM:SS
  lab <- gsub('^00h:', '', lab)              # Remove leading 00: if present
  lab <- gsub('^0', '', lab)                # Remove leading 0 if present
}
```

Including Plots

Density plot

```{r, echo=TRUE, message = FALSE,out.width="100%"}
ggplot(df, aes(x=msLapTime)) + 
 geom_histogram(aes(y= after_stat(density)), colour="black", fill="white", bins = 12)+
 geom_density(alpha=.2, fill="#FF6666") +
 xlab("Tempo da volta") + ylab("Densidade")
```

Scatter Plot

```{r,out.width="100%",message=FALSE}
plot <- ggplot(df, aes (x = LapNumber, y = msLapTime))+
  geom_point(alpha = 0.2) + 
  geom_smooth(method = lm,se = F)+ 
  scale_y_continuous(label=timeHMS_formatter, breaks = c(70000,80000,90000,100000,110000,120000)) +
  xlab("Número da volta") + ylab("Tempo da Volta")

plot
```

Scatter Plot grouped by Circuit as color

```{r,out.width="100%",message=FALSE}
plotCircuit <- ggplot(df, aes(x = LapNumber, y = msLapTime, color = Local))+
  geom_point(alpha = 0.2)+
  geom_smooth(method = lm,se = F) + 
  labs(color = "Grande Prêmio", xlab("Número da volta")) +
  scale_y_continuous(label=timeHMS_formatter, breaks = c(70000,80000,90000,100000,110000,120000)) + 
  xlab("Número da volta") + ylab("Tempo da Volta") +
  scale_color_discrete(name=NULL) +
  theme(legend.position="top", legend.box = "horizontal")
plotCircuit
```

Scatter Plot of each circuit separately

```{r,out.width="100%",message=FALSE}
plotWrapEventos <- ggplot(df, aes (x = LapNumber, y = msLapTime))+
  geom_point(alpha = 0.2) + 
  geom_smooth(method = lm,se = F)+ 
  scale_y_continuous(label=timeHMS_formatter) +
  facet_wrap(~ Local, ncol = 6) +
  xlab("Número da volta") + ylab("Tempo da Volta") 

plotWrapEventos
```
Behavior for each lap

Filtering Dutch Grand Prix for exemplification purposed.

```{r}
df_dutch = df[df[,"Local"] == "Holanda",]
```

Creating auxiliary data grouped by Driver (LastName) to plot stints separately.

```{r,out.width="100%",message=FALSE}
dataLine1 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line1 = mean(Stint_1))

dataLine2 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line2 = mean(Stint_2))

dataLine3 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line3 = mean(Stint_3))

dataLine4 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line4 = mean(Stint_4))

dataLine5 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line5 = mean(Stint_5))

dataLine6 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line6 = mean(Stint_6))

dataLine7 <- df_dutch %>%
  group_by(LastName) %>%
  summarize(line7 = max(LapNumber))
```

Plotting Dutch GP LapTimes for each driver, specifying Tyre and Stints.

```{r,out.width="100%",message=FALSE}
plotHolanda <- ggplot(df_dutch, aes (x = LapNumber, y = msLapTime, col = TyreType, linecolor = "red",group = 1))+
  geom_point(alpha = 0.2) + 
  geom_line() +
  geom_vline(data = dataLine1, aes(xintercept = line1), linetype="dashed", color = "red") +
  geom_vline(data = dataLine2, aes(xintercept = line2), linetype="dashed", color = "red") +
  geom_vline(data = dataLine3, aes(xintercept = line3), linetype="dashed", color = "red") +
  geom_vline(data = dataLine4, aes(xintercept = line4), linetype="dashed", color = "red") +
  geom_vline(data = dataLine5, aes(xintercept = line5), linetype="dashed", color = "red") +
  geom_vline(data = dataLine6, aes(xintercept = line6), linetype="dashed", color = "red") +
  geom_vline(data = dataLine7, aes(xintercept = line7), linetype="dashed", color = "red") +
  labs(color = "Tipo de pneu", linetype = "Tipo") +
  theme(legend.position="top") +
  scale_y_continuous(label=timeHMS_formatter) +
  xlab("Número da volta") + ylab("Tempo da Volta") + 
  facet_wrap(~ LastName, ncol = 5)

plotHolanda
```

Explain behaviour.

Boxplot: Laptimes for each circuit

```{r,out.width="100%",message=FALSE}
boxplotCircuits <- ggplot(df,aes(x = Local, y = msLapTime)) +
  scale_y_continuous(label=timeHMS_formatter, breaks = c(70000,80000,90000,100000,110000,120000)) +
  geom_boxplot() + 
  xlab("") + ylab("")+ 
    theme(axis.text=element_text(size=12))+
  guides(x =  guide_axis(angle = 45))
boxplotCircuits
```

Boxplot: Laptimes for each driver

```{r,out.width="100%",message=FALSE}
df_color = df[c("Team","TeamColor")]
df_color = unique(df_color)
df_color = df_color[order(df_color$Team),]
```

```{r,out.width="100%",message=FALSE}
boxplotDriver <- ggplot(df,aes(x = LastName, y = msLapTime,fill = Team)) +
  scale_y_continuous(label=timeHMS_formatter, breaks = c(70000,80000,90000,100000,110000,120000)) +
  geom_boxplot() + 
  scale_fill_manual(name = "Equipe", values = as.character(df_color$TeamColor)) +
  xlab("") + ylab("")+ 
  theme(legend.position = "top",
        legend.background = element_rect(fill = "transparent"),
        axis.text=element_text(size=12))+
  guides(x =  guide_axis(angle = 45), fill = guide_legend(direction = "horizontal"))

boxplotDriver
```
Fastest lap for each race

```{r,message=FALSE}
fastest_laps = data.frame()
for (event in unique(df$EventName)){
  x = df[df$EventName == event,] %>%
  filter(msLapTime == min(msLapTime)) %>% 
  select(EventName, DriverNumber, LastName, LapNumber, msLapTime)
fastest_laps = rbind(fastest_laps,x)
}
fastest_laps
```


